\documentclass[10pt,conference,draftclsnofoot,onecolumn]{IEEEtran}
\usepackage{listings}
\usepackage[dvipsnames]{xcolor}
\usepackage{color}
\usepackage{anysize}
\usepackage{hyperref}
\usepackage[backend=bibtex]{biblatex}
\usepackage{amsmath}
\marginsize{2cm}{2cm}{2cm}{2cm}
\addbibresource{bib.bib}

\lstdefinelanguage
   [x86Extended]{Assembler}     % add an "x86Extended dialect of Assembler
   [x86masm]{Assembler}         % based on the "x86masm" dialect
   % Define new keywords
   {morekeywords={rdrand, cpuid}}

\lstdefinestyle{customc}{
  belowcaptionskip=1\baselineskip,
  breaklines=true,
  frame=L,
  xleftmargin=\parindent,
  language=C,
  showstringspaces=false,
  basicstyle=\footnotesize\ttfamily,
  keywordstyle=\bfseries\color{OliveGreen},
  commentstyle=\itshape\color{Fuchsia},
  identifierstyle=\color{black},
  stringstyle=\color{Bittersweet},
}

\lstdefinestyle{customasm}{
  belowcaptionskip=1\baselineskip,
  frame=L,
  xleftmargin=\parindent,
  language=[x86masm]Assembler,
  basicstyle=\footnotesize\ttfamily,
  commentstyle=\itshape\color{Fuchsia},
}

\lstset{escapechar=@,style=customc}

\usepackage{graphicx}

\begin{document}

\begin{titlepage}
    \centering
    {\scshape\LARGE Oregon State University \par}
    \vspace{1cm}
    {\scshape\Large CS 444 Operating Systems II\par}
    \vspace{1.5cm}
    {\huge\bfseries Assignment I: Getting Acquainted\par}
    \vspace{2cm}
    {\Large\itshape Ian Kronquist\par}
    \vfill
    \par
    Professor~Kevin \textsc{McGrath}

    \vfill

% Bottom of the page
    {\large \today\par}
\end{titlepage}


\author{\IEEEauthorblockN{Ian Kronquist}
\IEEEauthorblockA{School of Electrical and\\Computer Science\\
Oregon State University\\
Corvallis, Oregon\\
kronquii@oregonstate.edu}}


\begin{abstract}
In this project I compiled version 3.14.26 of the Linux kernel on a Centos
server and ran it in a virtual machine using qemu. Additionally, I completed a
project implementing a classic problem in the field of concurrency.
\end{abstract}

\bigskip
\bigskip
\bigskip

\section{Log of Commands}
This is a selection of the commands I ran to clone and build the Linux kernel
on the os-class server. Not all of them were run in the same shell.
Note that I use several aliases for git, vim, and other common commands.
\begin{verbatim}
mkdir /scratch/spring2016/cs444-072/
cd /scratch/spring2016/cs444-072/
git init
mkdir ass1
v qemu.sh
chmod +x qemu.sh
./qemu.sh
touch core-image-lsb-sdk-qemux86.ext3
cp /scratch/opt/environment-setup-i586-poky-linux .
source environment-setup-i586-poky-linux
git clone git://git.yoctoproject.org/linux-yocto-3.14
cd linux-yocto-3.14
git checkout v3.14.26
git checkout -b mcgrath
cd ..
cp /spring2015/files/config-3.14.26-yocto-qemu .
ga config-3.14.26-yocto-qemu
git commit
ga qemu.sh
git commit
cp config-3.14.26-yocto-qemu linux-yocto-3.14/.config
cd linux-yocto-3.14
make -j4 all
./qemu.sh
v qemu.sh
./qemu.sh
gdb
killall qemu-system-i386
\end{verbatim}

\section{QEMU Flags}
These flags can be found in the QEMU man page \cite{qemu(1)}.
\begin{enumerate}
\item \texttt{-gdb tcp::5572} Open to connection with the GNU Debugger over TCP on port 5572. GDB can also use a custom protocol to communicate with qemu.
\item \texttt{-S} Do not immediately start the CPU when qemu starts.
\item \texttt{-nographic} Disable graphical output and redirect the emulated serial port to the console.
\item \texttt{-kernel bzImage-qemux86.bin} Specifies the image to boot. This can be any multi-boot compliant kernel.
\item \texttt{-drive file=core-image-lsb-sdk-qemux86.ext3,if=virtio} Define a new drive with the file \texttt{core-image-lsb-sdk-qemux86.ext3} over the virtio interface.
\item \texttt{-enable-kvm} Use KVM virtualization.
\item \texttt{-net none} Don't configure any network devices.
\item \texttt{-usb} Enable USB driver.
\item \texttt{-localtime} Use the system's local time.
\item \texttt{-no-reboot} Don't reboot when if kernel crashes. A common cause would be a Triple Fault. Triple faults are nasty.
\item \texttt{-append "root=/dev/vda rw console=ttyS0 debug"} Use the provided commands as the kernel command line.
\end{enumerate}

\section{What's the Point?}
There are two parts of this assignment which have separate goals. The point of the first part is to familiarize students with the process of building the Linux Kernel and running it using QEMU. The point of the concurrency portion of the assignment is to refresh students' memories about using the pthread APIs and introduce them to a famous concurrency problem.

\section{Personal Approach}
When I began the assignment I started by reviewing the man pages for the various pthread functions. I also contemplated the pros and cons of using a mutex or a semaphore for controlling access to the buffer. I decided that a mutex was more applicable in this case because semaphores are more useful when only a fixed number of threads can share a resource, while a mutex can only be held by one thread at a time. It also allowed me to stick to the pthread API without using the posix semaphore API as well. Next I defined the data structures for the project. I made sure to pick meaningful data types for the item structure within the bounds of the description of the assignment. In this case, a thread cannot sleep for a negative period of time, so using a signed integer for a waiting period is an unwise decision. Another example of thoughtful data type choice involves the \texttt{BufferFill} variable, which holds the level of how full the buffer is. This should be a \texttt{size\_t} which is a data type for holding the maximum size of an array on the target system as an unsigned integer type.

I noticed that in the assignment there were several ranges for different waiting periods. I decided that these periods were fairly arbitrary so I sequestered them into their own functions so they can be changed later without hunting through the rest of the code. This promotes a strong separation of concerns.

I also created a \texttt{get\_random\_number\_function} which delegated to a function pointer called \texttt{real\_get\_random\_number}. Changing the function pointer changes the pseudo-random number generator which is called. This design means that the runtime check which determines which generator to use only happens once when the generator is configured. An alternative approach would be to use a static local variable which is set the first time the \texttt{get\_random\_number\_function} function is called, but this would make the function slower the first time it is called, harder to debug, and if another PRNG implementation is ever added this function will have to be changed significantly in addition to changing the PRNG initialization function. This design promotes separation of concerns and future extensibility at the expense of an extra function call every time a random number must be retrieved and additional global state.

\lstset{language=C,caption={Dispatching the Appropriate Implementation Using a Function Pointer},label=getRandomNumber}
\begin{lstlisting}

static unsigned int (*real_get_random_number)(void);

unsigned int get_random_number() { return real_get_random_number(); }

void random_number_init()
{
	if (cpu_supports_rdrand()) {
		real_get_random_number = native_get_random_number;
	} else {
		initialize_mersenne_twister(time(0));
		real_get_random_number = mersenne_twister_get_random_number;
	}
}
\end{lstlisting}

I then created created a debug target in the makefile which defines an additional preprocessor symbol named DEBUG and adds several flags. I then added \texttt{\#ifdef DEBUG} sections around my debugging print statements. Additionally, I made the debug print statements output to stderr.

I made sure the minimize the critical section between where the mutex lock is acquired and where it is released. The space between these two calls should include as few statements as possible and these statements should only modify the \texttt{BufferFill} variable or the buffer itself. A thread should not sleep when a lock is being held. Additionally, getting a random number is an expensive operation so random numbers should not be generated when a lock is being held.

\lstset{language=C,caption={Minimizing Critical Section},label=getRandomNumber}
\begin{lstlisting}
		// Acquire the lock
		if (pthread_mutex_lock(&BufferLock)) {
			perror("Acquiring lock in Producer");
			exit(EXIT_FAILURE);
		}

		// Check to see if the buffer is full
		if (BufferFill == BUFFER_SIZE) {
			// The buffer is full. Give back the buffer.
			if (pthread_mutex_unlock(&BufferLock)) {
				perror("Releasing lock in Producer");
				exit(EXIT_FAILURE);
			}

			continue;
		}
		// Set the new item
		Buffer[BufferFill] = born;

#ifdef DEBUG
		fprintf(stderr, "Producing value %zu\n", BufferFill);
#endif
		// Indicate the item has been produced
		BufferFill++;
		// Release the lock
		if (pthread_mutex_unlock(&BufferLock)) {
			perror("Releasing lock in Producer");
			exit(EXIT_FAILURE);
		}

\end{lstlisting}

All calls to the pthread family of functions should have their errors checked. An error caused by any of these functions is fatal and should lead to the immediate termination of the thread. Calls to sleep, printf, or fprintf may fail without serious consequences to the program, and their errors are not checked.

The algorithm is incredibly simple. Three producers and three consumers are started by the main thread and after that they dive into a free-for-all as bloody as a rugby scrum to acquire the single lock. More efficient algorithms exist, but are not true to a legalistic reading of the assignment.

The assembly for the assignment was incredibly straightforward. The function \texttt{cpu\_supports\_rdrand} does not need to save any registers since the only registers modified are the caller save registers. The \texttt{cpuid} instruction is run, which modifies the eax, ecx, and edx registers \cite{2_intel}. Only the content of the edx register is applicable, and only one bit in that register is necessary to determine whether the rdrand instruction is supported. This bit is masked out and the value is placed in the eax register to be the return value.

\lstset{language=[x86Extended]Assembler,caption={Using the CPUID Instruction to Determine If the CPU Supports RDRAND},label=getRandomNumber}
\begin{lstlisting}
; uint32_t cpu_supports_rdrand()
global cpu_supports_rdrand
cpu_supports_rdrand:
	; Registers ECX, EAX, and EDX are caller saved
	cpuid
	mov edx, eax
	and eax, (1<<30)
	ret
\end{lstlisting}

\lstset{language=[x86Extended]Assembler,caption={Using the RDRAND Instruction to Generate Random Numbers},label=getRandomNumber}
The \texttt{native\_get\_random\_number} is ridiculously simple. It simply calls the rdrand instruction which places the random value in the eax register to be the return value \cite{1_intel}.
\begin{lstlisting}
; uint32_t native_get_random_number()
global native_get_random_number
native_get_random_number:
	rdrand eax
	ret
\end{lstlisting}

I encountered an interesting difference between how GCC links code on OS X and on Linux. On OS X with GCC 5, external symbols' names are prefixed with an underscore, but they are not on Linux. An OS X specific macro created an alias with the necessary underscore.

There is an old story about how Kernighan and Ritchie were writing UNIX before the days of source control, and they would periodically merge their code together. One day they did this and discovered that they had implemented a single simple UNIX command line utility independently, but their separate implementations were byte for byte identical. Those of us who are not programming gods are not nearly so consistent. In order to ensure that the assignment is compliant with the Linux Kernel Style Guide I ran the program \texttt{clang-format} on it with a custom configuration. \texttt{clang-format} rewrites the code to rigidly follow a style guide. I found this custom configuration in the LLVM documentation.

\section{Testing}
I started by manually make sure that the debugging information was informative and manually checked that the threads were sleeping for the proper amount of time in the right order for the right amount of time, and that items were being consumed in the right order. Sample output of the debug mode is included. Note that the buffer size has been constrained to 4 in the debug builds so that the affects of the producer running out of space in the buffer can be observed in a shorter period of time. This value can be adjusted by defining a custom \texttt{BUFFER\_SIZE} at compile time. I also wrote a simple unit test. The test is not terribly complex. It uses simple asserts to ensure that the lock is unlocked after every call and that the buffer contents are not disturbed. The tests are in \texttt{test-producer-consumer.c} and can be built using the \texttt{make test} target.


\begin{verbatim}
Producer sleeping for 6
Producer sleeping for 5
Producer sleeping for 6
Producing value 0
Producer sleeping for 6
Start consuming value 0
Consumer sleeping for 5
Producing value 0
Producer sleeping for 4
Start consuming value 0
Consumer sleeping for 2
Producing value 0
Producer sleeping for 4
Start consuming value 0
Consumer sleeping for 2
35126303
Finish consuming value 0
420483259
Finish consuming value 0
1506170311
Finish consuming value 0
Producing value 0
Producer sleeping for 4
Producing value 1
Producer sleeping for 4
Start consuming value 1
Consumer sleeping for 2
Start consuming value 0
Consumer sleeping for 5
Producing value 0
Producer sleeping for 4
Start consuming value 0
Consumer sleeping for 3
1513645288
Finish consuming value 1
917313663
Producing value 0
Finish consuming value 0
Producer sleeping for 5
Start consuming value 0
Consumer sleeping for 3
Producing value 0
Producer sleeping for 5
Start consuming value 0
Consumer sleeping for 5
1518140619
\end{verbatim}

\section{What I learned}
I spent a great deal of time learning more about how Latex can build nifty citations with bibtex. I also learned about some new Latex macros and how to build PDF documents without \texttt{pdflatex}. I got the opportunity to review the pthreads API. I also got learned a few new arguments for QEMU, which should come in handy because I use that tool so often. I was able to learn more about the Linux config file, which I knew existed but had never played with. The rest of the Linux build process is reassuringly boring. I also refreshed my understanding of the Mersenne Twister. Its implementation is beguilingly simple but theoretically dizzying.


\section{Work Log}
\begin{tabular}{|p{5cm}|p{5cm}|p{5cm}}
    \textbf{Approximate Start Time} & \textbf{Approximate Duration} & \textbf{Activity} \\
    \hline
    19:00 Tuesday, March 29 & 30 min & Copy Linux build environment and configuration files \\
    19:30 Tuesday, March 29 & 1 hour 15 min & Implement data structures, multi-threading, and native assembly PRNG \\
    16:30 wednesday, march 30 & 15 min & Build Linux and log in with gdb. \\
    17:00 wednesday, march 30 & 1 hour 15 min & Implement Mersenne Twister, cpuid, and runtime PRNG determination \\
    19:00 Wednesday, March 30 & 15 min & Clean up C code \\
    19:15 Wednesday, March 30 & 2 hours 20 min & Write report \\
    21:35 Wednesday, March 30 & 5 min & More C code clean up \\
    21:40 Wednesday, March 30 & 1 hour & Learn bibtex and write a shell script for creating git log \\
    23:10 Wednesday, March 30 & 45 min & Learn about code listings and add them to the report. Define a custom dialect of assembler using the Latex listing packages so cpuid and rdrand show up as instructions. \\
    23:30 Wednesday, March 30 & 10 min & Prepare report for submission, assuming that the proper packages will be installed on os-class \\
    8:45 Tuesday, April 5 & 15 min & Write test because McGrath explicitly asked for one in class despite not mentioning one in the assignment.
\end{tabular}

\bigskip
\bigskip

\section{Git Logs}
\begin{tabular}{l l l}\textbf{Detail} & \textbf{Author} & \textbf{Description}\\ \hline\href{git@github.com:iankronquist/cs444/commit/6c4c4ad85b24e29c66726d07a80b0d9b05791010}{6c4c4ad} & Ian Kronquist & Preserve env setup script for bash\\\hline
\href{git@github.com:iankronquist/cs444/commit/3b610803bb893ba4d4c2bf7fc0ed67306e884456}{3b61080} & Ian Kronquist & Add long qemu command in a script\\\hline
\href{git@github.com:iankronquist/cs444/commit/83c50214d9dd20bfd3cc722178af0a18bc4d1497}{83c5021} & Ian Kronquist & Add yocto build config\\\hline
\href{git@github.com:iankronquist/cs444/commit/9e3dd48a769717e235ec85308267d1b6db4e95bf}{9e3dd48} & Ian Kronquist & First pass as implementation of assignment 1\\\hline
\href{git@github.com:iankronquist/cs444/commit/517b86ff39f3a9e738eb940cf5833c2f0b97cb43}{517b86f} & Ian Kronquist & Remove vim swapfiles\\\hline
\href{git@github.com:iankronquist/cs444/commit/984b2540a9ed74f5042680ddedfc52038284e6d7}{984b254} & Ian Kronquist & Finished main part of the assignment\\\hline
\href{git@github.com:iankronquist/cs444/commit/5a593b9919d77a4e071d5b2397bb494003ce8a37}{5a593b9} & Ian Kronquist & Create Makefile which builds rdrand version\\\hline
\href{git@github.com:iankronquist/cs444/commit/25d0acf4229a54bf6bc5207be44677f41589cfc1}{25d0acf} & Ian Kronquist & Add debug target\\\hline
\href{git@github.com:iankronquist/cs444/commit/68c6230302be556de31b395f8c14b650b16b8039}{68c6230} & Ian Kronquist & Fix typo in rdrand version\\\hline
\href{git@github.com:iankronquist/cs444/commit/f6b831677fdf90ec9658466e6afe8fbd75d31177}{f6b8316} & Ian Kronquist & Fix debug print statements\\\hline
\href{git@github.com:iankronquist/cs444/commit/d85785b20039eac76880d1f86105e9f8d03e3895}{d85785b} & Ian Kronquist & Add a README\\\hline
\href{git@github.com:iankronquist/cs444/commit/5b429c1763541f159d23490d3b12166302cefc10}{5b429c1} & Ian Kronquist & Determine prng at runtime\\\hline
\href{git@github.com:iankronquist/cs444/commit/2e1d1d6fc6019e67e686e65e637c4c1ec495488f}{2e1d1d6} & Ian Kronquist & Link with -lpthread\\\hline
\href{git@github.com:iankronquist/cs444/commit/dcd7396df1b76690538705a49fdb7ee0edf8eee7}{dcd7396} & Ian Kronquist & Fix typo in register name\\\hline
\href{git@github.com:iankronquist/cs444/commit/4e0bdfcc3d106814ecd84b42e5234ec644f5f3f0}{4e0bdfc} & Ian Kronquist & Remove unnecessary initializer\\\hline
\href{git@github.com:iankronquist/cs444/commit/880d8ebe8ddfc4c4550e07c30ecb037fcb09570e}{880d8eb} & Ian Kronquist & Fix linking on OS X\\\hline
\href{git@github.com:iankronquist/cs444/commit/9e6331ff9ab92c498639b8834fbb6d4b5a4f7403}{9e6331f} & Ian Kronquist & Add Linux coding style-like .clang-format file\\\hline
\href{git@github.com:iankronquist/cs444/commit/dac8e44c9ac4dfc092dac067c8014b18c46326f1}{dac8e44} & Ian Kronquist & Run clang-format on C files\\\hline
\href{git@github.com:iankronquist/cs444/commit/800cc38c9370a30fec56c865f505d19a88de5c63}{800cc38} & Ian Kronquist & First draft of report for assignment 1\\\hline
\href{git@github.com:iankronquist/cs444/commit/f285be63fbc6a89bdc2c5db115f7e1f372ad41f3}{f285be6} & Ian Kronquist & Add latex building bits to Makefile\\\hline
\href{git@github.com:iankronquist/cs444/commit/db301c2e64d57a285039892bf3b050ce0b8b9134}{db301c2} & Ian Kronquist & Improve pthread usage\\\hline
\href{git@github.com:iankronquist/cs444/commit/471954de94723642247f08c71c04b862a85e648a}{471954d} & Ian Kronquist & Run clang-format on producer-consumer.c\\\hline
\href{git@github.com:iankronquist/cs444/commit/cefbcbf80b87afe068c91004e5825d88b639fb89}{cefbcbf} & Ian Kronquist & Add new sections and listings\\\hline
\href{git@github.com:iankronquist/cs444/commit/2acfdad69042df82fb1c2cc46fde841aa167b4a7}{2acfdad} & Ian Kronquist & Build git\_log.text as part of pdf target\\\hline
\href{git@github.com:iankronquist/cs444/commit/3bbfb59c0f65f3496a75ba8dfa45e923ca39fb88}{3bbfb59} & Ian Kronquist & Add script which generates changelog latex doc\\\hline
\href{git@github.com:iankronquist/cs444/commit/a4552398b16056270c2ceb4b2944c12d0b530470}{a455239} & Ian Kronquist & Change path to kernel image\\\hline
\href{git@github.com:iankronquist/cs444/commit/61f46b28903ed766bb4474c22bb43f07329e847a}{61f46b2} & Ian Kronquist & Vendor latex dependencies\\\hline
\href{git@github.com:iankronquist/cs444/commit/372766a3af5cebe7bcb0e19526e703ead5546e24}{372766a} & Ian Kronquist & Merge branch 'master' of https://github.com/iankronquist/cs444\\\hline
\href{git@github.com:iankronquist/cs444/commit/67f91af6cfce11662d5e2e5d79abef1fe25ba588}{67f91af} & Ian Kronquist & Include dependencies manually\\\hline
\href{git@github.com:iankronquist/cs444/commit/35012305f5307c9f4cbeda825f8ac55bf231a3c8}{3501230} & Ian Kronquist & More latex files\\\hline
\href{git@github.com:iankronquist/cs444/commit/355a1fc34c346abd2dd0e306bc62267eea41bdff}{355a1fc} & Ian Kronquist & Make sure latex table doesn't extend past the page\\\hline
\href{git@github.com:iankronquist/cs444/commit/6e1f40269ebc2fa5277338f092d784355cba9864}{6e1f402} & Ian Kronquist & Fix escaping in git log table generation script\\\hline
\href{git@github.com:iankronquist/cs444/commit/7b2ad566e48c9c5483a6105728765c15b7497051}{7b2ad56} & Ian Kronquist & Replace echo with printf\\\hline
\href{git@github.com:iankronquist/cs444/commit/495939bec91e9233f6151a616d0811425470235e}{495939b} & Ian Kronquist & Fix escaping in git log generation shell script\\\hline
\href{git@github.com:iankronquist/cs444/commit/8ca0181ad52664f27ee80fb2829e1702c4334cfe}{8ca0181} & Ian Kronquist & First version of encrypted block device module\\\hline
\href{git@github.com:iankronquist/cs444/commit/53b6fe7ae70719a16a0c6b4e50ec72dd582ff989}{53b6fe7} & Ian Kronquist & Miscellaneous notes\\\hline
\href{git@github.com:iankronquist/cs444/commit/4295f57ebb82a2167c116d54306ab041a4c55ffa}{4295f57} & Ian Kronquist & Add git ignore files\\\hline
\href{git@github.com:iankronquist/cs444/commit/fd8e2ff3420b4410d9bac99406076c794f2e25ab}{fd8e2ff} & Ian Kronquist & Refactor core components into another file\\\hline
\href{git@github.com:iankronquist/cs444/commit/d1d9a87e0ed0365bb0c830ba279a2cb0d8eb0192}{d1d9a87} & Ian Kronquist & Tests\\\hline
\end{tabular}


\clearpage
\printbibliography

\end{document}
\bibliography{bib.bib}
\bibliographystyle{IEEEtran}
